---
title: "Phenotypic Data Part I"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pheno}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

This script analyzes and prepares phenotype data for genetic mapping. Like any other phenotypic dataset there are some errors that we will go through and fix.

* Notes:

IDs 5053–5060 are missing values in the Geno field; assign "Program-Round-2".

The Geno value "P1-OG-Mcgregor" should be "OG-Mcgregor"—actual parent identity is irrelevant at this stage.

The ID column is numeric; append "-F2" to each entry (e.g., "1001-F2") to match genotype file formatting.

Phenotype IDs match genotype entries, though the latter include extra well/plate info (e.g., "1001-F2-Alpha-A1").

After visualization, remove Parents, F1s, and population 1633 to format the phenotype file for R/qtl.


## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(geneticMapR)
library(stringr)
library(dplyr)
library(ggdist)
library(ggplot2)
library(rlang)
library(ggpubr)
library(readxl)
```



## Data

```{r data, message=FALSE, warning=FALSE}
# Ordered Markers
url1 <- "https://raw.githubusercontent.com/vegaalfaro/geneticMapRFiles/refs/heads/main/phenotype/phenotypes-binary.csv"

# Download file
if (!file.exists("local_copy_phenotypes-binary.csv")) {
  download.file(url1, destfile = "local_copy_phenotypes-binary.csv")}

# Read
pheno <- read.csv("local_copy_phenotypes-binary.csv")

```

Feel free to explore the phenotype file. This file is more than we need because it contains data for 3 populations and we will be only working with one.

## Curation

```{r curation, message=FALSE, warning=FALSE}
#--- Data curation ----

# 1. Populate correctly the Geno columns for IDs 5053:5060
pheno$Geno[pheno$ID %in% 5053:5060] <- "Program-Round2"

# 2. Rename parents OGMacgregor and Program Round1
pheno$Geno <- str_remove(pheno$Geno, "P1-|P2-")

# 3. Change the ID column
pheno$IDs <- paste(pheno$ID, pheno$Gen, sep = "-")


# Most important variables
vars <- names(pheno)[c(9, 10, 16, 22, 23, 24, 26)]
response_vars <- vars #make a copy

# Print variables we'll be keeping or working with
vars
```



## Data Exploration

We begin by calculating basic statistics. The `Geno` column indicates the group to which each individual belongs. There are three groups with 107 or more individuals, these correspond to the F~2~ populations. In addition, the dataset includes entries for the parents and F~1s~ of these segregating F~2~ populations.

```{r exploration, warning=FALSE, message=FALSE}

# Means by the column genotype for one variable
means_LW <- pheno %>% 
  group_by(Geno) %>% 
  summarize(mean = mean(length_width_ratio),
            sd = sd(length_width_ratio),
            min = min(length_width_ratio),
            max = max(length_width_ratio),
            n = n()
            )
head(means_LW, n=10)
```


### Multiple traits

If you have multiple traits, below is a loop that would help you to estimate the statistics for each trait and save it in a list.

```{r}

# List of columns/traits to compute statistics
response_vars <- vars

# Create empty vector to store results 
response_var_stats <- list()

# Iterate for each column 
for (col in response_vars){
  # compute statistics 
  stats <- pheno %>%
    group_by(Geno) %>%
    summarize(across({{ col }}, list(
      mean = mean,
      sd = sd,
      min = min,
      max = max, 
      n = ~sum(!is.na(.))
    )),
    .groups = "drop")
  
  # Add to the results list
  response_var_stats[[col]] <- stats
  
}

print(response_var_stats)
```

## Subset

We are interested in `population 2`. There is more data here than we need so will subset it.

```{r, message=FALSE, warning=FALSE}
# Population 2
Pop2 <- pheno[pheno$Geno %in% c("1629-CylindricalxFlat-F2", 
                                "Cylindra", 
                                "F1-1629", 
                                "Mono"), ]
```

### Parents data

We can also subset the data for the F~2~ **parents**. Including the parents in phenotypic evaluations is useful for comparison and helps interpret how they rank relative to the progeny of the segregating populations. It’s generally a good practice to include parental data when screening F~2~ individuals or, at the very least, to collect some phenotypic data on the parents for reference.

```{r}
#--- Subset parents only----

# Subset parents
all_parents <- pheno[pheno$Gen %in% c("P"),]

# Subset parents of population 1
parents <- pheno[pheno$Geno %in% c("OG-Mcgregor", 
                                   "Program-Round1" 
),]

```

### Multiple populations

If you have multiple populations and traits combined in a single dataset, you can use code like the example below to compare them. In this case, I show data for all three populations and illustrate the approach using two traits as an example.

```{r, fig.height=4.3, fig.width=6.5, message=FALSE, warning=FALSE}

plots2 <- list()

for (col in response_vars) {
  plot <- ggplot(pheno %>% filter(Geno %in% c("1592-ConicalxRound-F2", 
                                              "1633-RoundxCylindrical-F2", 
                                              "1629-CylindricalxFlat-F2")), 
                 aes(x = Geno, 
                     y = !!sym(col), 
                     fill = Geno)) +
   
     geom_violin(alpha = 0.5, color = "#2C3E50") +  # Rain (density)
    geom_boxplot(width = 0.2, alpha = 0.7, color = "#2C3E50", outlier.shape = NA) +  # Cloud (boxplot)
    geom_jitter(width = 0.15, alpha = 0.6, size = 1.5, color = "#A680B8") +  # Drops (data points)
    labs(x = "Group", y = col, title = paste0("Raincloud Plot of ", col)) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(size = 13, face = "bold"),
      axis.text.x = element_text(size = 12, color = "#34495E"), 
      axis.text.y = element_text(size = 12, color = "#34495E"), 
      axis.title.x = element_text(size = 12.5, face = "bold"), 
      axis.title.y = element_text(size = 12.5, face = "bold"),
      panel.background = element_rect(fill = "#FAF3E0", color = NA), # Light beige background
      panel.grid.major = element_line(color = "#D5DBDB", size = 0.5),
      panel.grid.minor = element_blank(),
      legend.position = "none"
    ) + coord_flip()
  
  plots2[[col]] <- plot
}

print(plots2[vars][1:2])   # Run this
```

## Publication ready plots for parent differences

In my previous linkage mapping projects, I’ve found it useful to highlight differences among the parents (see, for example, this publication: https://doi.org/10.1093/g3journal/jkae041). In this case the parents are `Cylindra` and `Mono`, two table beet varieties  Below, I demonstrate how to create a high-quality, publication-ready plot that shows the differences among the parental types.

This code is designed to work for multiple traits but it would work for only one variable with minor refactoring. I show only 2 traits for example purposes.

```{r, fig.height=4.3, fig.width=6.5, message=FALSE, warning=FALSE}
# Population 2
response_vars <- vars

# Y-Axis labels

ylab <- c(expression(Biomass ~ (mm ^ 2)),
          "Maximum width (mm)",
          "Width 50 (mm)",
          "Length (mm)",
          "Length to width ratio (mm)",
          expression("Shoulder area" ~ (mm ^ 2)),
          "Tip angle (degrees)")

# Subset parents of population 2
parents2 <- pheno[pheno$Geno %in% c("Mono", 
                                   "Cylindra" 
),]


plots_pa2 <- list()  # Initialize the list

for (i in seq_along(vars)) {
  col <- vars[i]  # Extract column name
  
  plot <- ggplot(parents2,
                 aes(x = Geno, 
                     y = !!sym(col), 
                     fill = Geno)) +
    # Add half-eye distribution
    ggdist::stat_halfeye(
      adjust = 0.85,
      alpha = 0.85,
      width = 0.6,
      .width = 0,
      justification = -0.2,
      point_colour = NA
    ) +
    # Add boxplot
    geom_boxplot(
      width = 0.15,
      alpha = 0.4, 
      outlier.shape = NA,
      show.legend = FALSE
    ) +
    # Add points
    geom_point(
      aes(color = factor(Geno)),
      size = 1.5,
      alpha = 0.5, 
      position = position_jitter(seed = 1, width = 0.1),
      show.legend = FALSE # Remove legend for points
    ) +
    # Labels and aesthetics
    labs(x = "", 
         y = ylab[i], # Correctly indexing ylab
         title = "Population 2"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(size = 13),
      axis.text.x = element_text(size = 12, color = "#34495E"), 
      axis.text.y = element_text(size = 12, color = "#34495E"), 
      panel.background = element_rect(fill = "#FAF3E0", color = NA), #  background
      panel.grid.major = element_line(color = "#D5DBDB", size = 0.1),
      panel.grid.minor = element_blank(),
      legend.position = "none"
    ) + 
    coord_flip()
  
  plots_pa2[[col]] <- plot  # Store plot in the list
}

# Check stored plots
plots_pa2[1:2] # Run this
```




