---
title: "Phenotypic Data Part II"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pheno_part_2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro
This vignette will deal with how to properly and in a reproducible way work your phenotypic and genotypic data in a format compatible with R/qtl, a widely used package in linkage mapping.


## Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(geneticMapR)
library(stringr)
library(dplyr)
library(ggdist)
library(ggplot2)
library(rlang)
library(ggpubr)
library(readxl)
```

## Data
We continue with the data of our previous script.

```{r data, message=FALSE, warning=FALSE}
# Phenotype
url1 <- "https://raw.githubusercontent.com/vegaalfaro/geneticMapRFiles/refs/heads/main/phenotype/phenotypes-binary.csv"

# Download file
if (!file.exists("local_copy_phenotypes-binary.csv")) {
  download.file(url1, destfile = "local_copy_phenotypes-binary.csv")}

# Read
pheno <- read.csv("local_copy_phenotypes-binary.csv")

# Most important variables
vars <- names(pheno)[c(9, 10, 16, 22, 23, 24, 26)]
response_vars <- vars #make a copy

knitr::kable(head(pheno), caption = "Preview of the phenotype dataset")
```

In addition, we will need some meta-data available from the same data repository `geneticMapRFiles`.
The file ID contains the column `VCF_name` which will match the ID in the genotype matrix. The phenotype and the genotype files share the same individuals but the names are slightly different. This ID file, 1001-F2-Alpha-A1, 1002-F2-Alpha-B1, 1003-F2-Alpha-C1). They share parto of the name but we need to make them match.


```{r, message=FALSE, warning=FALSE}

url2 <- "https://github.com/vegaalfaro/geneticMapRFiles/raw/main/IDs/ID.csv"

# Download file
if (!file.exists("local_copy_ID.csv")) {
  download.file(url2, destfile = "local_copy_ID.csv")}

# Read
ID_Pedigrees <- read.csv("local_copy_ID.csv")

# Make as character
ID_Pedigrees$ID <- as.character(ID_Pedigrees$ID)

head(ID_Pedigrees)
knitr::kable(head(ID_Pedigrees), caption = "Preview of the ID Meta-data")
```

## Prepare data for R/QTL

This section filters the dataset to include only F~2~ individuals, excluding parents and F~1s~. It also matches phenotype IDs to their corresponding VCF IDs for integration with genotypic data, and selects only the relevant columns required for downstream analysis in R/qtl.

```{r}
# We subset like we did in the previous script.
Pop2 <- pheno[pheno$Geno %in% c("1629-CylindricalxFlat-F2", 
                                "Cylindra", 
                                "F1-1629", 
                                "Mono"), ]

# Keep only F2 Individuals
Pop2_progeny <- Pop2[Pop2$Gen == "F2",]
Pop2_progeny$ID <- as.character(Pop2_progeny$ID) # Declare ID as character

# Add VCF ID (ID that matches the geno file)
Pop2_progeny <- Pop2_progeny %>%
  left_join(ID_Pedigrees %>% select(ID, VCF_name), by = "ID")

# Ensure "VCF_name" is included in the selection
keep <- c(vars, "VCF_name")

# Select only the desired columns and rename VCF_name to ID
Pop2_progeny <- Pop2_progeny %>%
  ungroup() %>%  # Remove any existing grouping (Annoying dplyr thing)
  dplyr::select(all_of(keep)) %>%
  rename(ID = VCF_name)


knitr::kable(head(Pop2_progeny), caption = "Preview of the Phenotypic data")
```





