---
title: "Drawing a QTL trace"
output: rmarkdown::html_vignette
date: "2025-07-10"
vignette: >
  %\VignetteIndexEntry{qtl_trace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Intro

This vignette demonstrates how to prepare a QTL trace plot in a reproducible way and publication ready. The steps include loading data, extracting physical or genetic positions, identifying significant thresholds, and visualizing results with `geneticMapR` functions based on `ggplot2` and `ggpubr.`


```{r setup}
library(geneticMapR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(cowplot)
library(ggpubr)
```

## Load CIM Results
Let's load the previous results.

```{r data}
load("../data/plot_trace_pop2.RData")
```

## Setup

Using names(M2$pheno)[-8] dynamically selects all phenotype variables except the ID (8th column). 

```{r}
response_vars <- names(M2$pheno)[-8]  # Drop ID column

# Define the traits manually for clarity or fixed order
traits <- c("tip_angle_top", "width_50", "length", "shoulder_top", "length_width_ratio", "biomass")

```


## Helper functions

```{r}
combine_qtl <- function(results, trait_names) {
  do.call(rbind, lapply(names(results), function(i) {
    df <- results[[i]]
    df$response_var <- trait_names[match(i, names(results))]
    df
  }))
}

add_physical_pos <- function(df) {
  marker.name <- rownames(df)
  split_marker <- strsplit(marker.name, "[_.]")
  df$phys.pos <- as.numeric(sapply(split_marker, `[`, 2)) / 1e6
  df
}

get_thresholds <- function(qtl_df, perm_df, trait) {
  sig_thresh <- perm_df["5%", 1]
  qtl_df %>%
    filter(lod >= sig_thresh) %>%
    distinct(chr) %>%
    mutate(hline = sig_thresh, response_var = trait)
}
```

## Thresholds

```{r}
thresholds_df2 <- bind_rows(lapply(traits, function(trait) {
  get_thresholds(cim_qtl_results2[[trait]], cim_perm_pop2[[trait]], trait)
}))
```


## Trait Colors

```{r}
trait_colors <- c(
  tip_angle_top = "#d55e00",
  width_50 = "#cc79a7",
  length = "#0072b2",
  shoulder_top = "#f7a600",
  length_width_ratio = "#009e73",
  biomass = "black"
)

```



## Combine and Annotate QTL Results

```{r}
pop2 <- do.call(rbind, lapply(names(cim_qtl_results2), function(i) {
  df <- cim_qtl_results2[[i]]
  df$response_var <- response_vars[match(i, names(cim_qtl_results2))]
  df
}))

pop2 <- add_physical_pos(pop2)
```

## Known QTLs (Vertical Lines)
```{r}
vline_df2 <- data.frame(chr = unique(pop1$chr),
                        vline = c(NA, NA, NA, 48, NA, 16.6, 12.75, NA, NA))

```


```{r}
pop2_plot <- ggplot(pop2, aes(x = pos, y = lod, color = response_var)) +
  geom_line(linewidth = 0.85, alpha = 0.6) +
  geom_hline(data = thresholds_df2, aes(yintercept = hline, group = response_var),
             size = 1.05, alpha = 0.75) +
  geom_vline(data = vline_df2, aes(xintercept = vline),
             linetype = "dotted", color = "black", alpha = 0.85, size = 0.63) +
  facet_wrap(~chr, nrow = 1) +
  theme_test() +
  theme(legend.position = "none") +
  scale_color_manual(values = trait_colors) +
  labs(x = "Position (cM)", y = "LOD", color = "Trait") +
  scale_x_continuous(limits = c(0, 80)) +
  ggtitle("Population 2")

pop2_plot

```


