---
title: "Recode Markers"
output: rmarkdown::html_vignette
date: "2025-04-18"
author: "Andrey Vega Alfaro"
vignette: >
  %\VignetteIndexEntry{Recode Markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

This article demonstrates how to process, clean, and recode (phase) SNP markers from a VCF file to generate genotype matrices compatible with genetic mapping workflows. We will use helper functions from the `geneticMapR` package as well as dependencies such as `MapRtools`, `VariantAnnotation`, and `Rsamtools`. The example focuses on an F2 population, where genotype recoding is based on founder descent.

## Setup

We will need a series of packages, below we show how to install them if needed.


```{r setup, message=FALSE, warning=FALSE}
# Load or install packages if needed 
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager", quiet = TRUE)
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools", quiet = TRUE)

if (!requireNamespace("geneticMapR", quietly = TRUE)) devtools::install_github("vegaalfaro/geneticMapR", quiet = TRUE)
if (!requireNamespace("MapRtools", quietly = TRUE)) devtools::install_github("jendelman/MapRtools", quiet = TRUE)

if (!requireNamespace("VariantAnnotation", quietly = TRUE)) BiocManager::install("VariantAnnotation", quiet = TRUE)
if (!requireNamespace("Rsamtools", quietly = TRUE)) BiocManager::install("Rsamtools", quiet = TRUE)

# Load libraries
library(geneticMapR)
library(MapRtools)
library(VariantAnnotation)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(Rsamtools)
library(ggpubr)
```

## Load and Explore VCF File
The VCF file is somewhat large (19.8 MB) to be included as a system file. [geneticMapRFiles](https://github.com/vegaalfaro/geneticMapRFiles) is a data-only repository that will help us save data and run through examples without much issues. Let's get a local copy of the VCF file using `utils::download.file()`

```{r path, message=FALSE, warning=FALSE}
# Let's define our URL from the data repository that accompanies this R package
vcf_url <- "https://raw.githubusercontent.com/vegaalfaro/geneticMapRFiles/main/vcf/SNP_updated_IDs_sorted2.vcf.gz"

# Download file
if (!file.exists("local_copy.vcf.gz")) {
  download.file(vcf_url, destfile = "local_copy.vcf.gz")
}
```

The code below loads the vcf file using the function readVcf. 

```{r Load data}
# Let's read the vcf into our environment
vcf2 <- VariantAnnotation::readVcf("local_copy.vcf.gz", genome = "unknown")
```

Let's explore the vcf file. The VCF contains several fields including GT (genotype), DP (depth), GQ (genotype quality) and PL (sample-level annotations)

```{r vcf_overview}
# Overview
sapply(geno(vcf2), class)
```


## Extract Genotype and Coverage Information
Let's extract the genotype and depth which are saved in the GT and DP fields using VariantAnnotation's geno function.

```{r GT and DP, message=FALSE}
GT <- geno(vcf2)$GT
DP <- geno(vcf2)$DP
```

Once extracted, we can estimate depth statistics.

```{r DP stats}
summary(as.numeric(DP))
```


## Convert Genotype Call to Dosage

If we inspect the `GT` object,

```{r GT overview}
GT[1:3, 1:3]
```

genotypes are coded as `"0/0"`, `"0/1"`, or `"1/1"`, representing the maximum likelihood call: `0` for the reference (REF) allele and `1` for the alternate (ALT) allele.
For example: 

- `"0/0"` is a dosage of 0 ALT alleles, 
- `"0/1"` or `"1/0"` is a dosage of 1 ALT allele and 
- `"1/1"` corresponds to a dosage of 2 ALT alleles.

The `convert_to_dosage` [family of functions](https://vegaalfaro.github.io/geneticMapR/reference/index.html) in `geneticMapR` handles this conversion easily. For polyploids (e.g., `"0/0/0/0"`), check out `convert_to_dosage_advanced` and `convert_to_dosage_flex`.

Let's convert the genotype calls into dosages:

```{r convert to dosage}
geno_1629 <- convert_to_dosage_flex(GT)
```

Let's take a look at the conversion.
```{r dosage geno overview}
geno_1629[1:3, 1:3]
```

We have converted to dosage from allele format call.

## Filter Markers and Individuals by Missing Data

The function `filter_missing_geno()` takes a genotype matrix with rows representing genetic markers and columns representing individuals. The `geno_matrix` argument specifies the genotype matrix.

The `threshold` argument allows to specify the max proportion of missing data before an individual or marker is removed. Ranges from 0 to 1 and the default is 0.10.

Let's filter by `markers` first and accept 10% missing data.

```{r missing data filter}
result <- filter_missing_geno(geno_matrix = geno_1629, 
                              threshold = 0.10,
                              filter_by = "markers")

# Extract our result
filtered_geno <- result$filtered_geno
```

We can see that no marker was removed. Our markers have all less than 10% missing data. We can confirm it using a plot.

```{r plot missing data markers, fig.align='center',fig.alt = "plot missing", fig.height=4.3, fig.width=6.1, message=FALSE, warning=FALSE}
# Prepare data
missing_values <- base::as.data.frame(result$pct_missing)
colnames(missing_values) <- "missing"

# Visualize missingness
ggplot(missing_values, aes(x = missing)) +
  geom_histogram()

```

We can also filter by missing data per `individuals`.

```{r  missing data per individuals}
geno_missing_filtered <- filter_missing_geno(filtered_geno,
                                             threshold = 0.10,
                                             filter_by = "individuals")
filtered_geno <- geno_missing_filtered$filtered_geno
```

The function created a message letting us know that 7 individuals were removed as they have data that exceeds the 0.1 threshold.

Let's see an example on how to access the results.

```{r filter data}
# Access outputs
filtered_geno <- geno_missing_filtered$filtered_geno # Filtered geno matrix
missing_values <- base::as.data.frame(geno_missing_filtered$pct_missing) #
colnames(missing_values) <- "missing"

# Removed individuals
removed_individuals <- data.frame(ind = geno_missing_filtered$removed_individuals)

```

Let's see which individuals were removed

```{r ind removed, message=FALSE, warning=FALSE}
print(removed_individuals)
```

We can visualize how much missing data we still have in the kept individuals. It is no greater than 7% which is in agreement with our filtering parameters.

```{r plot missing data individuals, fig.height=4.3, fig.width=6.1, fig.align='center', warning=FALSE, message=FALSE, fig.alt="missing data plot"}
# Visualize missing data
ggplot(missing_values, aes(x = missing)) +
  geom_histogram()
```

## Visualize Genotype Data

We can also plot a histogram of genotypic values using `plot_genotype_histogram()`.

```{r viz genotype data, fig.align='center', fig.height=4.3, fig.width=6.1, warning=FALSE, message=FALSE}
plot_genotype_histogram(filtered_geno)
```

Let's also take a look at the coverage of the markers across the genome and chromosomes. This function is based on MapRtool's `plot_coverage.` I just added a few customization. Check it out [here](https://github.com/jendelman/MapRtools/tree/master/R). 

Because our filtered genotype matrix has the marker names in the row names in the format "CHR7_192222" we can use the function `extract_map()` from `geneticMapR` to easily create a physical map with columns "marker" "chrom" and "position".

The function `extract_map()` is very flexible but the row names should contain components (chromosome, position) separated by a symbol (e.g., _, -, .). Use arguments `chrom_index` and `pos_index` to specify their positions, and split_symbol to define the delimiter.

In our example **"CHR7_192222"**, `chrom_index = 1` as the chromosome is the first field before the delimiter  "_". If our markers were coded as **W257B_"CHR7_192222"** then `chrom_index = 2` and `pos_index = 3` as chromosome and position are in the second and third field respectively.

`extract_map` is tested to work with the symbols "_", "-" and ".". Other valid separators may also be valid

```{r extract map}
# Extract physical map
map <- extract_map(genotype_matrix = filtered_geno,
                   chrom_index = 1, # Index for Chromosome 
                   pos_index = 2, 
                    markers = FALSE,  # If TRUE, includes original marker names
                    split_symbol = "_") 
```

Let's see the first few lines of the map

```{r map head, message=FALSE, warning=FALSE}
# See map heading
head(map)
```


Now we are ready for plotting:

```{r plot coverage 1, fig.align='center', fig.width=6.1, fig.height=4.3, fig.alt="coverage plot"}
geneticMapR::plot_coverage(map=map, customize = TRUE)
```


## Genotype Frequency 

We can evaluate genotype frecuency using the `freq()` function. The function computes the relative frequency of each genotype (typically coded as 0, 1, or 2) across markers or individuals.

#### Genotype Frequency by Marker
We first calculate genotype frequencies by marker:

```{r freq 1}
geno.freq.mar <- freq(filtered_geno, input_format = "numeric", by = "markers")
```

This returns the relative frequency of each genotype (0, 1, 2) at each marker locus across all individuals, allowing identification of loci with high homozygosity or heterozygosity.

It can also help to filter out markers downstream.

Let's take a look
```{r freq head markers}
head(geno.freq.mar)
```


#### Genotype Frequency by Marker

To evaluate genotype frequncies for individuals across all markers:

```{r freq 2}
geno.freq.ind <- freq(filtered_geno, input_format = "numeric", by = "individuals")
```

This produces the relative frequency of each genotype for every individual, helping to pinpoint individuals with unusually high or low homozygosity levels.


Let's take a look
```{r freq head inds}
head(geno.freq.ind)
```

## Marker Type Analysis

Markers can be classified based on the genotypes of two parents:

- **Homozygous**: Fixed for either one allele (P1 = 0 & P2 = 2 or P1 = 2 & P2 = 0:)
- **Non-polymorphic**: Both parents have the same genotype (e.g., P1 = P2 = 0, 1, or 2)
- **Heterozygous**: One or both parents are heterozygous (e.g., P1 = 1 & P2 = 0)

Non-polymorphic markers should always be removed. Heterozygous markers can optionally be excluded depending on your analysis goals.

To filter out heterozygous markers and keep only homozygous markers, we can use `filter_geno_by_parents()`.

```{r filter geno by parents}
# Declare the name of the parents as they appear in the column of your geno
P1 <- "P2550-Cylindra-P1-Theta-A9"
P2 <- "P2493-Mono-P2-Theta-B9"

# Run our function to fiter out heterozygous markers
geno_homozygous <- filter_geno_by_parents(filtered_geno, P1, P2)
```

`P1` and `P2` are the column names for the parental genotypes. `filter_geno_by_parents` retains only homozygous polymorphic markers between two specified parents. 

**Note**: The `recode_markers() `function also performs this filtering automatically. The user can specify also if `recode_markers()`should keep het markers or not. Let's see.

  
## Recode Genotypes

`geneticMapR::recode_markers()` is the heart of the package. This powerful function phases or recodes genotype marker data based on two parental references (arguments `parent1` and `parent2`).

Recoding markers is one of the key steps in genetic mapping. The markers are coded in a way such that the haplotypes coming from either parent can be tracked in the F~2~ progeny. 

```{r recode fig, echo=FALSE, fig.height=3.7, fig.width=10, message=FALSE, warning=FALSE}
library(ggpubr)
library(png)       # Needed for readPNG
library(grid)      # Needed for rasterGrob if you want to manipulate image directly
library(ggplot2)   # Needed for ggplot and theme

img1 <- readPNG("../man/figures/recode_full.png")
im_A <- ggplot() + 
    background_image(img1) 
    # This ensures that the image leaves some space at the edges
  
ggarrange(im_A)
```


### Typical representation

In the figure above panel A is a typical representation of an intercross to derive an  F~2~ mapping population. The parental lines are usually shown as fully homozygous:

- P1.a carries *only* the reference allele (blue, dosage = 0).
- P2.b carries *only* the alternate allele (orange, dosage = 2).

The resulting F1 is heterozygous (green, dosage = 1) at all loci.

F~2~ individuals show a segregating mix of genotypes (0, 1, 2). 

This representation is valid when one your parents is the *reference genome used to call SNPs*. If that is the case then no recoding needs to occur. However, that is a rather rare case. 

### Alternative representation

In this representation, parental genotypes, are assumed to be homozygous but are fixed for different alleles across loci, allowing for more complex and realistic representations:

- P1.b and P2.b carry *both* homozygous dosage values (0, or 2). Still homozygous but fixed for different alleles.

### Outcrossing species

Most often and in outcrossing species like carrot and table beets or others with high heterozygosity, the (inbred) parents may have mixed genotypes with not totally homozygous genotypes.
This is because some loci are lethal on either homozygous configuration.

- P1.c and P2.c carry *some* heterozygous markers (1) in a low proportion, with homozygous dosage values of 0 or 2.

 F~1s~ are mostly heterozygous, but due to genotyping error or residual heterozygosity some dosages of 0 and 2 are typically observed in these individuals


## Recode function

`recode_markers()` was created to help with the task of recoding the markers so that they reflect a the "typical representation". This helps the user track the origin of the haplotypes in the F~2~. **Essential for genetic mapping.**  The function is quite flexible and its main functionality is:

- Drop markers where either of the parent is NA
- Remove non-polymorphic markers automatically (both parents have the same genotype)
- If `handle_het_markers` = TRUE, allows parental heterozygous markers to be kept.

**Note**: there are different types of Het markers. Please see the [function documentation](https://vegaalfaro.github.io/geneticMapR/reference/index.html) for more information.


```{r recode}
# Retains only homozygous markers
hom_phased_geno_1629 <- geneticMapR::recode_markers(geno = filtered_geno,
                                parent1 = P1,
                                parent2 = P2,
                                numeric_output = TRUE,
                                handle_het_markers = FALSE)

# Allows for heterozygous markers
het_phased_geno_1629 <- geneticMapR::recode_markers(geno = filtered_geno,
                               parent1 = P1,
                               parent2 = P2,
                               numeric_output = TRUE,
                               handle_het_markers = TRUE)
```

Below is an example of how `recode_markers()` works. Is based on simulated data for convenience. 

```{r}
# Load the example dataset
data("simulated_geno")

# Check markers previous to recoding
print(simulated_geno)

# Recode the markers using the recode_markers() function
phased <- geneticMapR::recode_markers(simulated_geno, parent1 = "Parent1", parent2 = "Parent2")

# Print the output
print(phased)
```


## Tradeoff Between Marker Type and Genome Coverage

In outcrossing species I have noticed a tradeoff between coverage and the quantity of heterozygous markers.

Including heterozygous markers increases genome coverage compared to using only homozygous markers, but comes with tradeoffs. While heterozygous markers can improve coverage, especially in regions near the centromere, loci with segregation distortion or deleterious alleles, they may complicate downstream linkage group estimation due to their different segregation patterns.

The following section illustrates this tradeoff between marker type and genome coverage.

```{r tradeoff marker type v coverage, fig.align='center', fig.height=8, fig.width=8, warning=FALSE, message=FALSE}
a <- plot_coverage(map = extract_map(hom_phased_geno_1629)) + ggtitle("Homozygous")
b <- plot_coverage(map = extract_map(het_phased_geno_1629)) + ggtitle("Homozygous + heterozygous")
c <- plot_genotype_histogram(hom_phased_geno_1629)
d <- plot_genotype_histogram(het_phased_geno_1629)

ggarrange(a, b, c, d, 
          nrow = 2,
          ncol = 2
          )
```

  
## Genotype Frequencies and Filtering

`geneticMapR` includes functions to estimate frequencies like `freq()` in an easy way. The function allows to estimate genotype frequencies for later filtering.

```{r hom v het plot, warning=FALSE, message=FALSE, fig.align='center',  fig.height=8, fig.width=6}

# Get Frequencies
geno.freq.hom <- freq(hom_phased_geno_1629,  input_format = "numeric", by ="markers") 
geno.freq.het <- freq(het_phased_geno_1629, input_format = "numeric", by = "markers")

# Plots
het_plot <- frequency_plot(geno.freq.het) + ggtitle("Het markers, unfiltered")
hom_plot <- frequency_plot(geno.freq.hom) + ggtitle("Hom markers, unfiltered")

# Arrange
ggarrange(hom_plot,
          het_plot,
          nrow = 2)
```

From the plots above we see that there is an abundance of markers at frequencies of 0 and 1 across all gentoypes (i.e., 0, 1, 2).
We can use the function `filter_geno_by_freq()` to filter markers out by maximum genotype frequency and keeping a range of heterozygous frequencies, say no lower than 0.1 or hiher than 0.70. Which is reasonable for an F~2~ mapping populations.

The resulting plot shows a filtered marker set and the overabundance of markers at frequencies of 0 and 1 have been filtered out.

```{r freq filtering, fig.align='center', fig.height=4.3, fig.width=6.1}
# Heterozygous marker curation
het_phased_geno_1629_filt <- filter_geno_by_freq(het_phased_geno_1629, 
                                                 max_geno_freq = 0.90, 
                                                 het_freq_range = c(0.1, 0.70) 
)

# Create a dataset of frequencies based on the filtered data
geno.freq.het.filtered <- freq(het_phased_geno_1629_filt, 
                               input_format = "numeric", 
                               by ="markers") 


# Create plot
frequency_plot(geno.freq.het.filtered) + ggtitle("Het markers, filtered")

```

Below we see that our filtering paramaters have left us with a typical 1:2:1 segregation ratio even for markers that included heterozygous markers that were properly filtered.

```{r filtering comparison, fig.height=8.0, fig.width=6.0, message=FALSE, warning=FALSE}
v <- plot_genotype_histogram(het_phased_geno_1629_filt) +
  ggtitle("Homozygous + filtered het markers")

w <-  plot_genotype_histogram(hom_phased_geno_1629) +
  ggtitle("Homomozygous markers")

ggarrange(v, w,
          nrow = 2)

```


## Save Results

```{r results save, eval=FALSE}
save(hom_phased_geno_1629,
     het_phased_geno_1629,
     het_phased_geno_1629_filt,
     file = "processed_data/R_data/filtered_geno_matrices_1629.RData")
```


## Conclusion
A lot of genetic map construction revolves around the type of markers you have and how you make good use of them. Very many filtering steps are taken specially using GBS data to make sure the curated set of markers are markers you can trust and informative. 

This vignette provides a workflow for preparing and phasing genotypic data for genetic mapping analyses in an F~2~ population using tools from `geneticMapR` and other bioinformatics packages. Let's continue in the next article the genetic map construction.

## Session Information

```{r session }
sessionInfo()
```

